### 2015-02-27 icfp15-servo/open.tex
Some more polish

\paragraph{Just-in-time code} is produced from all modern JavaScript engines in order to
dynamically produce native code that is intended to execute more efficiently than an
interpreted strategy.
Unfortunately, this area is a large source of security bugs.
These bugs come from two sources.
First, there are potential correctness issues.
Many of these optimizations are only valid when certain conditions of the calling
code and environment hold, and ensuring the specialized code is called only when those
conditions hold is non-trivial to verify.
Second, dynamically producing and compiling native code and patching it into memory
while respecting all of the invariants required by the JavaScript runtime (e.g., the
garbage collector's tracing hooks or free vs. in-use registers) is also a challenge.

\paragraph{JavaScript object rooting} is a problem related to objects from the JavaScript
heap that are held onto by \Cplusplus object --- and sometimes vice-versa.
Today, all modern engines manually manage the boundary code to handle tracing
these objects, and this leads to leaking objects or, worse, freeing objects that
are still in use.
The lack of a typical garbage collector for these objects also means that cycles
between \Cplusplus and JavaScript need to be broken through ad-hoc means.

>>>>

\paragraph{Just-in-time code.} JavaScript engines dynamically produce native code that is
intended to execute more efficiently than an interpreted strategy.
Unfortunately, this area is a large source of security bugs.
These bugs come from two sources.
First, there are potential correctness issues.
Many of these optimizations are only valid when certain conditions of the calling
code and environment hold, and ensuring the specialized code is called only when those
conditions hold is non-trivial to verify.
Second, dynamically producing and compiling native code and patching it into memory
while respecting all of the invariants required by the JavaScript runtime (e.g., the
garbage collector's tracing hooks or free vs. in-use registers) is also a challenge.

\paragraph{JavaScript object rooting.} Objects from the JavaScript
heap  are held onto by \Cplusplus{} object --- and sometimes vice-versa.
Today, all modern engines manually manage the boundary code to handle tracing
these objects, and this leads to leaking objects or, worse, freeing objects that
are still in use.
The lack of a typical garbage collector for these objects also means that cycles
between \Cplusplus{} and JavaScript need to be broken through ad-hoc means.

\paragraph{Integer overflow/underflow.} It is still an open problem
to provide optimized code that checks for overflow or underflow without
incurring significant performance penalties.
The current plan for Rust is to have debug-only checking of integer ranges
and for Servo to run debug builds against a test suite, but that may miss
scenarios that only occur in optimized builds.

### 2015-02-27 icfp15-servo/rust.tex
Complete restructure
Some more polish

\section{Open}
\label{sec:open}

>>>>

\section{Rust}
\label{sec:rust}

TODO: basic overview, covering systems language basis; borrowing/ownership \& region/MLKit connection; ADT \& pattern matching; traits \& generics

\subsection{Concurrency and parallelism}

Rust contains library functionality for concurrency and data parallelism.
Concurrency, in particular, integrates cleanly with Rust's ownership types.
In safe concurrent programs, the data operated on by distinct threads is also itself distinct.
Through the use of Rust's ownership model, data cannot be owned by two threads at the same time.
For example, the following code generates a static error from the compiler because after the first
thread is spawned, the ownership of \lstinline{data} has been transferred into the closure associated
with that thread and is unavailable for transfer in to the closure for the second thread.
%http://is.gd/tkcD9R
\begin{lstlisting}
fn main() {
  let mut data = Box::new(0);

  Thread::spawn(move || {
    *data = *data + 1;
  });
  Thread::spawn(move || {
    *data = *data + 1;
  });
  print!(``{}'', data);
}
\end{lstlisting}

Data parallelism is a bit more tricky today because there is no way to \eg{} partition a vector in safe code
for a data parallel operation and then stitch it back together afterwards without performing a copy of the data.

### 2015-02-27 icfp15-servo/servo.tex
Some more polish

\section{Servo}

\subsection{Syntactic features}

Rust provides struct and enum types (similar to Standard ML's record types and datatypes~\cite{sml97-definition}) as
well as pattern matching.
These types and associated language features provided two large benefits to Servo over traditional browsers
written in \Cplusplus{}.
First, providing new abstractions and intermediate representations is syntactically easy, so there is very little
pressure to tack additional fields into classes simply to avoid creating a large number of new header and implementation
files.
More importantly, this style of programming that uses pattern matching over concrete types instead of
virtual function calls on a class hierarchy provides a non-trivial performance gain.
Virtual functions can both have an in-memory storage cost associated with the virtual fuction tables (sometimes many thousands of bytes\footnote{\url{https://chromium.googlesource.com/chromium/blink/+/c048c5c7c2578274d82faf96e9ebda4c55e428da}}) but more importantly
incur indirect function call costs.
All C++ browser implementations transform performance-critical code to either use the \lstinline[language=C]{final}
specifier wherever possible or specialize the code in some other way to avoid this cost.

Rust also attempted to stay close to familiar syntax, but did not require full fidelity or easy porting of
programs from languages such as \Cplusplus.
This approach has worked well for Rust because it has prevented some of the complexity that arose in Cyclone
with their attempts to build a safe language that preserved most C idioms.

@@

\subsection{Integer overflow/underflow}

\subsection{Affine types}
\label{sec:affine}

@@

\subsection{Unsafe code}
\label{sec:safety}

There are several categories of code in Servo that cannot be written in the safe subset of Rust.

\subsection{FFI calls}

Interoperating with native C code often requires raw access to memory allocated in Rust or returns
pointers to memory that was allocated by the C libraries.
Translations to an from these raw pointers requires the use of unsafe code.

@@

Rust's ownership model assumes that there is a single owner for each piece of data.
However, many data structures do not follow that model, in order to provide asymptotic
improvements in access.
For example, a doubly-linked list contains a back pointer to the previous element to aid
in traversals in the opposite direction.
Many optimized hashtable implementations also have both hash-based access to items
and a linked list of all of the keys or values.
\subsection{Libraries and abstractions}
\label{sec:libraries}

@@

\subsection{Concurrency and parallelism}

Concurrency and data parallelism are two places where Rust's type system is very close to perfectly suited.
In concurrent and (safe) data parallel libraries, the data operated on by distinct threads is also itself distinct.
Through the use of Rust's ownership model, data cannot be owned by two threads at the same time.
In the case of concurrency this works quite well --- there are often individual objects that need to be owned by
a single object at a time.

%http://is.gd/tkcD9R
\begin{lstlisting}
fn main() {
  let mut data = Box::new(0);

  Thread::spawn(move || {
    *data = *data + 1;
  });
  Thread::spawn(move || {
    *data = *data + 1;
  });
  print!(``{}'', data);
}
\end{lstlisting}

Data parallelism is a bit more tricky today because there is no way to \eg{} partition a vector in safe code
for a data parallel operation and then stitch it back together afterwards without performing a copy of the data.

TODO: code sample

>>>>

\section{Servo}
\label{sec:servo}
In this section, we cover specific areas of Servo's design or implementation that make use of Rust
and the impacts and limitations of these features.

\subsection{Rust's syntax}
Rust provides struct and enum types (similar to Standard ML's record types and datatypes~\cite{sml97-definition}) as
well as pattern matching.
These types and associated language features provided two large benefits to Servo over traditional browsers
written in \Cplusplus{}.
First, providing new abstractions and intermediate representations is syntactically easy, so there is very little
pressure to tack additional fields into classes simply to avoid creating a large number of new header and implementation
files.
More importantly, this style of programming that uses pattern matching over concrete types instead of
virtual function calls on a class hierarchy provides a non-trivial performance gain.
Virtual functions can both have an in-memory storage cost associated with the virtual fuction tables (sometimes many thousands of bytes\footnote{\url{https://chromium.googlesource.com/chromium/blink/+/c048c5c7c2578274d82faf96e9ebda4c55e428da}}) but more importantly
incur indirect function call costs.
All C++ browser implementations transform performance-critical code to either use the \lstinline[language=C]{final}
specifier wherever possible or specialize the code in some other way to avoid this cost.

Rust also attempted to stay close to familiar syntax, but did not require full fidelity or easy porting of
programs from languages such as \Cplusplus{}.
This approach has worked well for Rust because it has prevented some of the complexity that arose in Cyclone
with their attempts to build a safe language that preserved most C idioms.

@@

@@

That said, interoperating with native C code often requires raw access to memory allocated in Rust or returns
pointers to memory that was allocated by the C libraries.
Translations to and from these raw pointers requires the use of unsafe code.

@@

Rust's ownership model assumes that there is a single owner for each piece of data.
However, many data structures do not follow that model, in order to provide asymptotic
improvements in access.
For example, a doubly-linked list contains a back pointer to the previous element to aid
in traversals in the opposite direction.
Many optimized hashtable implementations also have both hash-based access to items
and a linked list of all of the keys or values.

\subsection{Libraries and abstractions}

@@
